---
title: "UTSW Nanocourse - Data Science in R : Fundamentals Exercise"
author: "Austin Marckx"
date: "2024-09-05"
output: html_document
---

```{r setup, include=FALSE}
library(ggpubr)
library(rstatix)
library(datasauRus)
library(GGally)
library(titanic)
library(randomForest)
library(caret)
library(MASS)
library(car)
library(tidyverse)
set.seed(1234)

# Use set dplyr fxns as priority for conflicts
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
```

# Section 1 : File I/O

## Ex 1.1: Load mtcars from the datasets module in R

Ensure the dataset is saved as a tibble

```{r}
# Load the library with the datasets
library(datasets)
#mtcars1 <- mtcars

# assign the mtcars dataset to variable in active environment
mtcars1 <-tibble(datasets::mtcars)

# print dataset to console
print(mtcars1)
```


## Ex 1.2: Correctly load mtcars from the file path provided:

```{r}
# DO NOT CHANGE: This is the relative file path to the dataset.
mtcars_filepath_ex1pt2 <- "data/cars.tsv"

# Load the dataset from the file; 
mtcars2 <- read_tsv(mtcars_filepath_ex1pt2)
mtcars2

```

## Ex 1.3: Correctly load mtcars from a filepath with atypical structure
```{r}
# DO NOT CHANGE: This is the relative file path to the dataset.
mtcars_filepath_ex1pt3 <- "data/cars_ex1pt3.tsv"

# Load the dataset from the file; 
mtcars3 <- read_tsv(mtcars_filepath_ex1pt3)
mtcars3
```

# Section 2: Knowing your data

```{r}
# The data is loaded into the environment for you
knowingYourData_Filepath <- "data/kyd.csv"
cars <- read_csv(knowingYourData_Filepath)
```


meta-data: A data frame with 32 observations on 6 variables.

| Column Index| Name |  Description | 
|-------------|------|--------------|
|[, 1] | mpg |	Miles/(US) gallon
|[, 2]|	cyl	|Number of cylinders
|[, 3]|	wt	|Weight (1000 lbs)
|[, 4]|	qsec	|1/4 mile time
|[, 5]|	am	| Transmission (0 = automatic, 1 = manual)
|[, 6]|	gear	|Number of forward gears
|[, 7]| rec   |Public vehicle recall : TRUE = Vehicle recalled o/w Not recalled


## Ex 2.1: First steps with new dataset 

```{r}
# Look at the head of the dataset
cars %>% head()

# Print the summary statistics
cars %>% summary()

# Check for missing values:
print(any(is.na(cars)))

# Extract rows with missing values
cars[!complete.cases(cars),]


```

## Ex 2.2: Describe the missing in `cars`, can you use the meta-data to make sure there are no missing values?

```{r}
# Fix Missing Data:
cars$rec <- ifelse(is.na(cars$rec), FALSE, TRUE)
```
## Ex 2.3: Column Structure and Datatypes

```{r}
# 1) Look at the structure:
cars %>% str() 

# 2) Transform the datatype of each column to what you think it should be based on the provided meta-data
cars <- cars%>% transform(
  cyl= as.factor(cyl),
  am = as.factor(am),
  gear = as.factor(gear)
)

```

## Ex 2.4: Exploratory Plotting

```{r, warning= FALSE, message = FALSE, fig.height = 9, fig.width = 9}
# Make a pairs plot and color it by the number of gears 
cars%>%
  ggpairs(aes(color=gear), progress = FALSE)
```

Can you identify any trends data?



# Section 3: Essentials of Data Wrangling

## Ex 3.1 Arranging, Selecting, and Dropping columns
```{r}
# From the dataset in section 2: 
# 1) rearrange the columns such that the gear comes first

# 2) drop the recall column 

# 3) save this in a variable called "newcars"
newcars <- cars%>% dplyr::select(gear, everything(), -rec) 
newcars

# Add in a new column `id` which assigns a unique number to each row
## HINT: This column will be important later
newcars <- newcars%>% mutate(id= row_number()) %>% dplyr::select(id, everything())
newcars

```


## Ex 3.2 : Filtering & subsetting rows

```{r}
# Filter the `newcars` to only include rows which whose cylinder value equal to the highest number of cylinders in the dataset. Save this variable as: `filtered_newcars`
filtered_newcars <- newcars%>% filter(cyl %in% max(as.numeric(levels(cyl))))
filtered_newcars

```
 
## Ex 3.3 : Group by, summarize, and mutate
 
```{r}
# Make a character vector with strings containing your two favorite colors
colors <- c("black", "blue")

# create a new column which randomly assigns one of the colors to each car in `filtered_newcars`
filtered_newcars<- filtered_newcars%>%
  mutate(paint= sample(colors, nrow(filtered_newcars),replace=TRUE))%>%
  transform(paint = as.factor(paint))
filtered_newcars

# Summarize the average weight for each color of car. Ensure this value is in lbs and round it to 2 decimal places. 
filtered_newcars%>%
  group_by(paint)%>%
  summarize(avg_color_wt= round(mean(wt*1000), 2))


```

## Ex 3.4 : Dataset Joins

```{r}
# Save a subset of `filtered_newcars` containing only the `id` and `paint` columns
cars_subset <- filtered_newcars %>% dplyr::select(id, paint)
cars_subset

# Now perform a left join with between `newcars` and `cars_subset`
cars_leftjoin <- newcars %>% left_join(cars_subset, by = 'id')
cars_leftjoin
```

## Ex 3.5 : More Dataset Joins
```{r}
# Save a subset of `filtered_newcars` containing only `id`, `paint`, and `cyl` columns
cars_subset <- filtered_newcars%>% dplyr::select(id, cyl, paint)
cars_subset
# Now perform a left join with between `newcars` and `cars_subset`
cars_leftjoin <- newcars%>% inner_join(cars_subset, by= c('id', 'cyl'))

# Validate that the inner_join dataset is equivalent to `filtered_newcars`
# Understanding when you can break up your data into smaller pieces is essential when working with large datasets

```
